---
title: "PA1_template"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(knitr)
```

## R Markdown document for the Assigmnet

### Step 1 : Load & preprocess data  
```{r load }
knitr::opts_chunk$set(cache = TRUE)
dat <- read.csv("activity.csv")
# Turn the date data into a valid date class
# Allows for easier processing
# Dates are in YYYY-MM-DD format
#dates <- strptime(dat$date, "%Y-%m-%d")
#dat$date <- dates
```

### Step 2 : Histogram of the total number of steps taken each day
```{r stpes_per_day}
# Unique dates in the data
unique_dates <- unique(dat$date)
#Unique intervals in the data 
unique_int <- unique(dat$interval)
# calculating steps per day
steps_date <- aggregate(steps ~ date,data = dat,sum)
#assigning column names to new data frame
colnames(steps_date) <- c("date","steps")
#Plotting required histogram
ggplot(steps_date,aes(x=steps))+geom_histogram(fill = "blue",binwidth = 1000)+labs(x = "No of Steps per Day",y ="Frquency",title = "Histogram of Total Steps taken per day")

```





### Step 3 :Mean and median number of steps taken each day

```{r mean_median}
mean_step <- mean(steps_date$steps,na.rm = TRUE)
median_step <- median(steps_date$steps,na.rm = TRUE)
print(mean_step)

```
Mean Steps per Day : `r mean_step`


Median Steps per Day : `r median_step`





### Step4 & 5:Time series plot of the average number of steps taken per interval of 5 min
```{r }
steps_interval <- aggregate(steps ~ interval,data = dat,FUN = mean,na.rm = TRUE)

plot(steps ~ interval,data = steps_interval,type = "l",xlab = "Time Intervals (5-minute)", ylab = "Mean number of steps taken (all Days)", main = "Average number of steps Taken at 5 minute Intervals",  col = "green")
max_interval <- steps_interval[which.max(steps_interval$steps),]
```
Interval With Max Steps : `r max_interval$interval` and No of Steps is `r max_interval$steps`





### Step6:Code to describe and show a strategy for imputing missing data
```{r impute}
#calculate no og missing values in dataset
num_na_dat <-  sum(num_na <- is.na(dat$steps))
num_na_dat

#The strategy that will be used to fill in the missing values in the data set is to replace all NA values with the mean of #that particular 5-minute interval the observation falls on

## This function returns the mean steps for a given interval
getMeanStepsPerInterval <- function(interval){
    steps_interval[steps_interval$interval==interval,"steps"]
    }
#Creating  a new dataset  but with the missing data filled in
complete_dat <-dat
## Filling the missing values with the mean for that 5-minute interval
flag = 0
for (i in 1:nrow(complete_dat)) 
        {
            if (is.na(complete_dat[i,"steps"]))
                {
                     complete_dat[i,"steps"] <- getMeanStepsPerInterval(complete_dat[i,"interval"])
        flag = flag + 1
                }
        }
 
#Check if there are any more NA values in complete_dat
num_na_comp_dat <-  sum(num_na <- is.na(complete_dat$steps))
num_na_comp_dat
```





###Step7:Histogram of the total number of steps taken each day after missing values are imputed
```{r stpes_per_day_comp_data}
# Unique dates in the data

# calculating steps per day
steps_date_comp <- aggregate(steps ~ date,data = complete_dat,sum)
#assigning column names to new data frame
colnames(steps_date_comp) <- c("date","steps")
#Plotting required histogram
ggplot(steps_date_comp,aes(x=steps))+geom_histogram(fill = "blue",binwidth = 1000)+labs(x = "No of Steps per Day",y ="Frquency",title = "Histogram of Total Steps taken per day")

steps_date_comp <- aggregate(steps ~ date,data = steps_date_comp,sum)
mean_step_comp <- mean(steps_date_comp$steps)
median_step_comp <- median (steps_date_comp$steps)

```
Before Imputation :
Mean Steps per Day : `r mean_step`   Median Steps per Day : `r median_step`

After Imputation :
Mean Steps per Day : `r mean_step_comp`   Median Steps per Day : `r median_step_comp`





###Step8:Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r Weekday_weekend}
## Finding out weekdays and weekends from complete data set and creating a new column for day of the week

complete_dat$days <-  weekdays(as.Date(complete_dat$date),abbreviate = FALSE)
# Agrregating the data 



## Seperating out Weekday and weekend activity 
 complete_dat_weekday <- subset(complete_dat, days != c("Sunday","Saturday") )
 complete_dat_weekend <- subset(complete_dat, days == c("Sunday","Saturday") )
 
## Averaging out weeekday and weekend activity
 complete_dat_weekday_avg <- aggregate(steps ~ interval,complete_dat_weekday,sum)
complete_dat_weekend_avg <- aggregate(steps ~ interval,complete_dat_weekend,sum)

# Creating a plot of weekdat vs weekend activity  
 par(mfrow = c(2,1))

with(complete_dat_weekday_avg,plot(interval,steps,xlab = "Interval", ylab ="Steps",type = "l",main = "Weekdays",col= "green"))
with(complete_dat_weekend_avg,plot(interval,steps,xlab = "Interval", ylab ="Steps",type = "l",main = "Weekends", col = "blue"))

```

